A Handbook for Tenured Track Faculty in Computer Science in the US

### 2024-08-18 demystify.tex
updates

\paragraph{Low Stipend?} Students often complain their stipend is too low. However, in most cases it is not bad, and you can live comfortably with it.  It might also be enough to support your spouse and kids (many CS PhD students have their families with them). So don't worry too much about the stipend.  If you're admitted to a good CS PhD program, you will be fine. A good school would know that it has to be competitive to attract students.  For example, at Mason, every year we discuss improving the benefits, and especially stipend, for our graduate students.

@@

\begin{commentbox}[Vu:]
  TA and RA at Mason have similar benefits in tuition waiving and insurance.  The college and department will set the rate for a 9-month graduate assistant stipend.  TA, which is paid by the department, will likely be that amount but RA might be higher depending on the stage of the student (1st year vs ABD\footnote{All but dissertation: close to graduate.}) and the prof.
  \tcblower
  Having health insurance is required at many US universities.  Do not assume that you're young and healthy and ignore insurance.  At Mason, and most CS PhD programs, your GTA or GRA comes with full insurance. In fact, at Mason your spouse/children will get a significant discount rate for health insurance.  So you will never have to worry much about health issues for you or your family here.
\end{commentbox}

@@

In addition to general fellowships that are available to all students, there are also specific fellowships, e.g., for US citizens and permanent residents, for underrepresented minority (URM) students (\autoref{sec:urm}), and students from specific countries. There are many fellowships available, so you should look for them and apply.

@@

\section{Funding In the Summer}\label{sec:summer-funding}

@@

@@

You \emph{do not} need to do a PhD in CS to work in your area of interest. For example, in addition to a traditional CS department, Mason has IST and Cybersecurity departments, both of which have faculty with PhD in CS and work on CS topics (e.g., AI, Security, Robotics).  So you still can do CS research and publish in CS-focused venues even if you're not in a traditional CS program.  It is common to see faculty with PhD in CS in a non-CS department as well as faculty with non-CS PhD in a CS department.

@@

Fortunately, while some non-US programs require finding an adviser and research topic before starting the PhD (\autoref{sec:non-us-differences}), CS PhD programs in the US will typically give you a couple of years to search for advisers and research topics.  This is especially true if you're admitted with TA (\autoref{sec:ta}), which gives you time to explore and find an adviser.

@@

Here is some advice for finding a prof.  Assuming you're not familiar with any particular one, then first search for profs. that share similar research interests.  For example, in CSRankings, if you want to work with PL, you can search for those published in PL conferences.  If you want to work with SE \emph{and} AI, you can search for faculty who work in both SE and AI.  Then you research that prof. and contact them (\autoref{sec:contact}).

@@

After all reviewers have evaluated an application, adcom chair will review all evaluations, look at entered notes, and ask reviewers to discuss and resolve discrepancies to reach a consensus (e.g., a reviewer wants to accept but the other wants to reject).  Typically, the decision is made entirely by the reviewers; there is no involvement from the adcom chair, department chair, or others. In most cases adcom members, even those reviewing the same application, make decisions independently and do not talk to each other (just a common practice to avoid biasing). In some rare cases we might (\autoref{sec:adcom-discuss}).

@@

If you're reaching out to a professor for the first time,  address them as Prof. or Dr. Lastname. Many international students use Prof. or Dr. FirstName LastName, but this can come across as if you're simply copying and pasting names. It's not necessary, so stick with Prof. or Dr. Lastname.

Furthermore, avoid using Mr., Mrs., Miss, or the professor's first name if you're not acquainted with them yet.  As you become more familiar with your prof and depending on their preferences, you may transition to addressing them by their first name.
For example, I prefer that my students and colleagues call me Vu. Some students call me \emph{Dr. Vu}, which I find a bit amusing but am totally fine with it.

@@

Note that in some universities the formal title Dr. Lastname is preferred over Prof. Moreover, be aware that not all faculty members may hold a PhD, in which case using Prof. Lastname is a suitable alternative. 	You just need to observe and follow the conventions at your particular institution.One way to determine how to address a professor is to observe how they sign their emails or how they introduce themselves in class. For example, I introduce myself as ``Vu'' in class.

\paragraph{Referring to professors you know} Because you are already familiar with these individuals, you can just informally use their names if they are OK with it as mentioned above (or Dr./Prof., if you want to be formal). You can also include their institution if it makes it more precise.  For example, I can say:  \emph{``I did my postdoc with Jeff Foster at Univ. of Maryland''}.

@@

\section{How much do \emph{I} cost?}\label{sec:ra-cost}

@@

@@

\section{Relationships}
There's a misconception that in the US it's all business, with professors as bosses who pay students for their work and that lab mates are just work colleagues; and that doing nice things means expecting something in return.

@@

>>>>

\paragraph{Low Stipend?} Students sometimes complain about their stipend being low. However, in most cases it is not bad, and you can live comfortably with it.  It might also be enough to support your spouse and kids (many CS PhD students have their families with them). So don't worry too much about the stipend.  If you're admitted to a good CS PhD program, you will be fine. A good school would know that it has to be competitive to attract students.  For example, at Mason, every year we try to improve the benefits, and especially stipend, for our graduate students.

@@

\begin{commentbox}[Vu:]
  TA and RA at Mason have similar benefits in tuition waiving and insurance.  The college and department will set the rate for a 9-month graduate assistant stipend.  TA, which is paid by the department, will likely be that amount but RA might be higher depending on the stage of the student (1st year vs ABD\footnote{All but dissertation: close to graduate.}) and the prof.
  \tcblower
  Having health insurance is required at many US universities.  Do not assume that you're young and healthy and ignore insurance (\autoref{sec:cultural-misc}).  At Mason, and most CS PhD programs, your GTA or GRA comes with full insurance. In fact, at Mason your spouse/children will get a significant discount rate for health insurance.  So you will never have to worry much about health issues for you or your family here.
\end{commentbox}

@@

In addition to general fellowships that are available to all students, there are also specific fellowships, e.g., for US citizens and permanent residents (\autoref{sec:domestic-students}), for underrepresented minority (URM) students (\autoref{sec:urm}), and students from specific countries. There are many fellowships available, so you should look for them and apply.

@@

\section{Miscs}

\subsection{Funding In the Summer}\label{sec:summer-funding}

@@

\subsection{How much do \emph{I} cost?}\label{sec:ra-cost}

@@

You \emph{do not} need to do a PhD in CS to work in CS areas. For example, in addition to a traditional CS department, Mason has IST and Cybersecurity departments, both of which have faculty with PhD in CS and work on CS topics (e.g., AI, Security, Robotics).  So you still can do CS research and publish in CS-focused venues even if you're not in a traditional CS program.  It is common to see faculty with PhD in CS in a non-CS department as well as faculty with non-CS PhD in a CS department.

@@

Fortunately, while some non-US programs require finding an adviser and research topic before starting the PhD (\autoref{sec:non-us-differences}), CS PhD programs in the US will typically give you a couple of years to ``shop'' for advisers and research topics.  This is especially true if you're admitted with TA (\autoref{sec:ta}), which gives you time to explore and find an adviser.

@@

Here is some advice for finding a prof.  Assuming you're not familiar with any particular one, then first search for profs. that share similar research interests.  For example, in CSRankings, if you want to work with PL, you can search for those published in PL conferences.  If you want to work with SE \emph{and} AI, you can search for faculty who work in both SE and AI.  Then you research about that prof. (e.g., go to their website) and contact them (\autoref{sec:contact}).

@@

After reviewers have evaluated an application, adcom chair will review all evaluations, look at entered notes, and ask reviewers to discuss and resolve discrepancies to reach a consensus (e.g., a reviewer wants to accept but the other wants to reject).  Typically, the decision is made entirely by the reviewers; there is no involvement from the adcom chair, department chair, or others. In most cases adcom members, even those reviewing the same application, make decisions independently and do not talk to each other (just a common practice to avoid biasing). In some rare cases we might (\autoref{sec:adcom-discuss}).

@@

If you're reaching out to a professor for the first time,  address them as Prof. or Dr. Lastname (if they do not have a PhD, then use Prof.). Many international students use Prof. or Dr. FirstName LastName, but this can come across as if you're simply copying and pasting names. So just stick with Prof. or Dr. Lastname.  Using Prof. is generally the safest option.

Furthermore, do not use Mr., Mrs., Ms, or Miss. This happens rarely, but I have seen  new students (e.g., undergraduate freshmen in the US) sometimes use these, which are used in K-12 schools but not in higher education.

Moreover, do not call the prof. by their first name at first.  As you become more familiar with your prof and depending on their preferences, you may transition to addressing them by their first name.
For example, I prefer that my students and colleagues call me Vu. Some students call me \emph{Dr. Vu}, which I find a bit amusing but am totally fine with it.

@@

Note that in some universities the formal title Dr. Lastname is preferred over Prof. Moreover, be aware that not all faculty members may hold a PhD, in which case using Prof. Lastname is a suitable alternative. 	You just need to observe and follow the conventions at your particular institution. One way to determine how to address a professor is to observe how they sign their emails or how they introduce themselves in class. For example, I introduce myself as ``Vu'' in class.

\paragraph{Referring to professors you know} When referring or talking about a prof (e.g., your mentor) that you know, you can just informally use their names if they are OK with it as mentioned above (or Dr./Prof., if you want to be formal). You can also include their institution if it makes it more precise.  For example, I can say:  \emph{``I did my postdoc with Jeff Foster at Univ. of Maryland''}.

@@

@@

\chapter{Glossary}\label{sec:glossary}

\begin{description}

  \item[ABD] All But Dissertation. A PhD candidate who has finished all course work and exams and only needs to write and defend their dissertation.

  \item[Adcom] Admission Committee. The group of faculty members who review applications and make admission decisions.
  \item [Adcom members] Faculty members who are part of the admission committee. These are the people who review applications and make admission decisions.
  \item[Adcom chair] The faculty member who leads the admission committee. This person is often not involved in individual admission decisions but oversees the process (e.g., resolve disputes, ensure fairness).
  \item[Advisor/Supervisor]: A faculty member who guides and mentors a PhD student throughout their research. This person typically plays a crucial role in a student's academic journey. In the US, the term ``adviser'' is more commonly used than ``supervisor''.
  \item[Cohort] A group of students who start a program at the same time and take classes together. You should get to know your cohort as they will be your colleagues and friends during your PhD.
  \item  [Deferred Admission] An option allowing admitted students to postpone their start date, typically by one year.
  \item [Diversity Statement] A document sometimes required in applications where applicants describe their contributions to diversity and inclusion, or how their background adds to the diversity of the academic community. This document is often required for faculty positions, but some universities also require it for graduate admissions.
  \item[Domestic Student] A student who is typically a US citizen or permanent resident (``greencard holder'').
        But sometimes it can also refer to students who did their undergrad (or MS) at a US university.
  \item[Field or Area] A specific research area, e.g., Machine Learning, Computer Vision, Software Engineering, etc.
  \item [In-state vs. Out-of-state] In-state tuition is the tuition rate for students who are residents of the state where the university is located. Out-of-state tuition is the tuition rate for students who are not residents of that state. In-state tuition is typically much lower than out-of-state tuition. PhD students typically do not have to pay tuition as it is covered by their funding, but this might be important for MS students.
  \item[LoR] Letter of Recommendation. A letter written by a professor or supervisor that assesses your qualifications and potential for graduate study (\autoref{sec:lor}).
  \item[SOP] Statement of Purpose. A document written by yourself to explains your research interests, background, and reasons for applying to a PhD program (\autoref{sec:research-statement}).

  \item[RA] Research Assistantship. A form of funding where you work as a research assistant for a professor.
  \item[TA] Teaching Assistantship. A form of funding where you work as a teaching assistant for a professor.
  \item[TOEFL] Test of English as a Foreign Language. A standardized test to measure the English language ability of non-native speakers.
  \item[IELTS] International English Language Testing System. A standardized test to measure the English language ability of non-native speakers.
  \item[GRE] Graduate Record Examination. A standardized test that is an admissions requirement for many graduate schools in the US. However, it is not required for most CS PhD programs (\autoref{sec:standard-tests}).

  \item [Lab] Research Lab. A group of researchers (profs. and students) working on a common research area, e.g., a Software Engineering Lab.  Typically a lab is led by a single professor, but sometimes it can be co-led by multiple professors. For example, my lab is DynaROARS, which focuses on my own research area (software verification), but I am also part of the more general SWE Lab at Mason with other faculty in Software Engineering.

  \item[MS] Master of Science. A graduate degree that typically requires 2 years of study.
  \item [PI]  Principal Investigator. A lead researcher on a grant or research project, e.g., a PI on an NSF grant. Prospective students often use this term to refer to a professor they are interested in working with.  While this term is used frequently in online forums among students, we do not use it this way in academia, e.g., ``who is your PI?'' might only make sense to fellow Reddit applicants, but might not be understood by others, ``who is your adviser?'' is what you want to use.
    \item[STEM] Science, Technology, Engineering, and Mathematics. A term used to group together these academic disciplines. CS is considered a STEM field.
  \item[URM] Underrepresented Minority. Refers to groups that are underrepresented in higher education, e.g., Black, Hispanic, and Native American students.
  \item[NSF] National Science Foundation. A US government agency that supports research and education in science and engineering. NSF is a major source of funding for RAs in CS research. There are other agencies that support RAs but they might require US citizenship or permanent residency.
        \item [R1 vs. R2 Universities] Research 1 (R1) universities are universities with the highest level of research activity. Research 2 (R2) universities have a high level of research activity but not as high as R1 universities.  R1 universities are typically more prestigious and have more funding and resources for research.  R2 universities are also good but might have fewer resources and funding.
  \item [Rolling Admission] A process where applications are reviewed as they are received, and decisions are made throughout the admission cycle.  This is in contrast to a process where all applications are reviewed after the deadline and decisions are made at the same time. Rolling admission is more common for MS programs and less common for PhD programs (e.g., at Mason, MS is rolling and PhD is not).
  \item [Top-tier conferences and journals] These are the most prestigious venues for publishing research in CS.  For example, in Software Engineering, top-tier conferences include ICSE, FSE, and ASE and top-tier journals include TSE and TOSEM.  Publishing in these venues is highly competitive and prestigious, and can significantly improve your chances of getting admitted to a good PhD program.

        \item [Stipend] A fixed regular sum paid to students as part of their funding package (i.e., salary). This is typically paid monthly or bi-weekly.  Stipend is typically paid to PhD students and not MS students.
  \item[REU] Research Experience for Undergraduates. A program funded by the NSF to provide research opportunities for undergraduate students.
    \item[OPT] Optional Practical Training. A program that allows F-1 students to work in the US for up to 12 months after graduation.
  \item[CPT] Curricular Practical Training. A program that allows F-1 students to participate in internships or practical training that is an integral part of their academic curriculum.
  \item[F-1] Student Visa. A visa that allows international students to study full-time at an accredited institution in the US.
  \item[F-2] Dependent Visa. A visa that allows the spouse and children of an F-1 visa holder to live in the US.
\end{description}

@@

\section{Maintaining Good Relationships with your Profs.}
There's a misconception that in the US it's all business, with professors as bosses who pay students for their work and that lab mates are just work colleagues; and that doing nice things means expecting something in return.

@@

  \section{Miscs}\label{sec:cultural-misc}
  Several other common surprises for international students (and foreigners in general) in the US:

  \paragraph{Healthcare System}  You (and your spouse) will need health insurance! Otherwise you will be charged a lot for healthcare services. However, as mentioned in \autoref{sec:funding}, your  TA/RA  will cover health insurance (and your spouse/children also get insurance or discount for it). However, be aware that heathcare services can still be expensive even with insurance. So  know what your insurance covers and be prepared for unexpected costs.  Moreover, healthcare system has many confusing jargons such as HMO, PPO, deductibles, copays, and coinsurance.  Get yourself familiar with them and ask for help when needed.

  \paragraph{Tipping Culture} Unlike many other countries, tipping is expected for various services, especially in restaurant. So adding an extra 15--20\% to your bills is common, especially in restaurants.
  You should also tip for other services, e.g., Uber, taxi, haircuts, and hotel services.  The minimum wage for tipped employees is lower than the standard minimum wage, so tips are an important part of their income.  If you do not tip, it is considered rude and disrespectful.

  \paragraph{Car Dependency} Most of the US is highly car-dependent. If you do not have a car, you will need to rely on friends, Uber, or public transportation, which can be inconvenient and time-consuming. Most international students end up getting a driver's license and buying a car.

	%\paragraph{Individualism} The strong culture of individualism in the US can be both liberating and isolating. People are more focused on their personal rights and tend to be less concerned with what others think.

  % \paragraph{Racism and Segregation} While some find the US to be less racist than expected, others note that racism and segregation are still prevalent, often manifesting subtly in daily life or visibly in segregated neighborhoods.

	%\paragraph{Litigiousness} The US is known for its litigious culture, where lawsuits are common, and people can be sued over seemingly minor issues. This can be surprising for newcomers used to different legal systems.

  % \paragraph{Consumerism and Scale} The US is often associated with large portions, big cars, and massive stores like Costco. The emphasis on consumption and abundance is notable.

	% \paragraph{Religious Influence} Religion, particularly Christianity, plays a more visible role in public and political life in the US compared to many other countries.

### 2024-08-18 faculty.tex
updates

Depending on your research area, you might focus on publishing at top conferences or journal.  You probably should know what's popular in your area already (if not, ask your senior colleagues). If you're at an R1 university, you will want to publish at top venues, and not just any venue.  You should also aim to publish at a variety of venues, and not just one.  This will show that your research is of interest to a wide range of people.

@@

\subsection{Short papers, workshops, etc} Short papers in workshop or non-research tracks at a conference (e.g., tool, demo, vision, etc) do not count as much.
They describe work that might not be as mature (e.g., new idea) or as complete (e.g., a tool or a demo) as a full paper, and also have higher acceptance rate. Many people use them as a way to get feedback on their work before submitting to a full paper venue.  Most new professors do not focus on these kinds of papers. Many P\&T committees do not count or care much about these and having too many of them can be a negative signal.

Personally, I find that such papers are very useful as preliminary work for writing grant proposals. They show that you have a new idea and good preliminary results that are worth pursuing further. Sometimes a full paper is not as good because the research is relatively complete with strong results, which bring the concern that you already did all the cool work and the proposal is minor and not exciting.

@@

\subsection{Your university does not recognize conference value} For most non-CS engineering fields journal are more prestigious than conferences. Places with a strong CS department are well aware of the value of top confs., but many other places are not (especially smaller universities or CS depts run by engineering people). These places also do not know much about the quality of the venue, and so in many cases put a workshop paper in the same class as a top conference paper and also cannot differentiate between a good vs bad journals in CS (because they do not know CS well). Also,  they often value quantity over quality because they cannot properly evaluate paper quality.

@@

\chapter{Obtaining Funding}\label{chap}
Without a doubt, grants are one of the most, if not \emph{the} most, important factors in P\&T (or any kind of performance) evaluation.  The most common reason for not getting tenure is the lack of funding. Ideally research results, e.g., publications (\autoref{chap:research}), would be the more important thing for a researcher; but in reality, funding is often considered more important for tenure evaluation at most R1 universities.
The amount of funding you have is a direct reflection of your research impact and productivity.  It is also a reflection of your ability to write proposals, manage projects, and work with collaborators. More realistically, grants are needed to pay for your students, travel conference, research equipments, and even your own summer salary.

@@

\paragraph{Personal Experiences}
As with most tenure track professors, I focus mostly on NSF grants and most of my funding comes from NSF.  I only have one grant from Army Research Office (ARO) in which I am a co-PI and didn't have to write the proposal. I have tried to get other DoD funding, but without much success, I found that DoD program managers are very picky and often have their own agenda.  I have also received internal funding from my university and industrial gifts from Amazon, and Facebook.  I have found that industrial gifts are the best kind of funding, as they are unrestricted and have no associated timeline by when you should spend them.  They are also highly competitive and prestigious, and are a great way to show research impact.

>>>>

Depending on your field, you might focus on publishing at top conferences or journal.  You probably should know what's popular in your area already (if not, ask your senior colleagues). If you're at an R1 university, you will want to publish at top venues, and not just any venue.  You should also aim to publish at a variety of venues, and not just one.  This will show that your research is of interest to a wide range of people.

@@

\subsection{Short papers, workshops, etc} Short papers in workshop or non-research tracks at a conference (e.g., tool, demo, vision, etc) do not count as much.
They describe work that might not be as mature (e.g., new idea) or as complete (e.g., a tool or a demo), and also have higher acceptance rate. Many people use them as a way to get feedback on their work before submitting to a full paper venue.  Many P\&T committees do not count or care much about these and having too many of them can be a negative signal.

While many faculty do not value short papers, I find that they are are very useful for writing grant proposals. They show that you have a new idea and good preliminary results that are worth pursuing further. Sometimes a full paper is not as good because the research is relatively complete with strong results, which bring the concern that you already did all the cool work and the proposal is minor and not exciting.

@@

\subsection{Your university does not recognize conference value}

Journals, not conferences, are the main venue for publication for most non-CS engineering fields. While places with a strong CS department are well aware of the value of top confs. in CS, many other places are not (especially smaller universities or CS depts run by engineering people). In many cases such places put a workshop paper in the same class as a top conference paper and also cannot differentiate between a good vs bad journals in CS (because they do not know CS well). Also,  they often value quantity over quality because they cannot properly evaluate paper quality.

@@

\chapter{Obtaining Funding}\label{chap}
Without a doubt, grants are one of the most, if not \emph{the} most, important factors in P\&T (or any kind of performance) evaluation.  The most common reason for not getting tenure is the lack of funding. Ideally research results, e.g., publications (\autoref{chap:research}), would be the more important thing for a researcher; but in reality, funding is often considered more important for P\&T at most R1 universities.
The amount of funding you have is a direct reflection of your research impact and productivity. More practically, grants are needed to pay for your students, travel conference, research equipments, and even your own summer salary.

@@

\paragraph{Personal Experiences}
As with most TT professors, I focus mostly on NSF grants.  I only have one grant from Army Research Office (ARO) in which I am a co-PI and didn't have to write the proposal. I have tried to get other DoD funding, but without much success, I find that DoD program managers difficult to make contact with and to get feedback from.  I have also received internal funding from my university and industrial gifts from Amazon, and Facebook.  I have found that industrial gifts are the best kind of funding, as they are unrestricted and have no associated timeline by when you should spend them.  They are also highly competitive and prestigious, and are a great way to show research impact.
